---
title: "S&P 500 Index Times Series Analysis"
author: "Ashrita Kodali"
date: "2025-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary of File:

We will be using R to conduct the time series analysis. In this case, the tidyverse
and dplyr packagtes will first be used to organize the initial dataset and obtain
average monthly returns. Once the data is processed, the astsa package will be used to
assess both the assumptions of the ARIMA and GARCH models by creating and assessing
times series plots and ACF and PACF plots. We will then train both the ARIMA and GARCH
models (using the astsa and fGarch packages respectively) and assess their respective
model assumptions. Once the models are built, the performance of these models will be
then assessed using metrics such as AIC, BIC, MAE, MSE, and RMSE. These values will then
be formatted and organized using kableExtra (it is important to note that the file
must be knitted into an HTML in order to view these figures properly). 

# Data Preprocessing:
```{r, message = FALSE, warning = FALSE}
df <- read.csv("stock_data.csv", header = TRUE)

# load in libraries
library(tidyverse) # used for data cleaning
library(dplyr) # used for data cleaning
library(astsa) # used for ACF + PACF plots and ARIMA model building
library(fGarch) # used for GARCH model building
library(kableExtra) # used to organize and create tables

# data cleaning
df <- select(df, date, return)
df <- drop_na(df)

df$date <- as.Date(df$date, format = "%m/%d/%y")
df$month <- as.numeric(format(df$date, format = "%m"))
df$year <- as.numeric(format(df$date, format = "%Y"))

data <- df %>% group_by(year, month) %>%
  summarize(avg_return = mean(return))

data <- as.data.frame(data)

avg_returns <- data$avg_return

returns <- ts(avg_returns, frequency = 12, start = c(2015, 1))

# separating data into training and testing
training <- returns[1:108]
testing <- returns[109:122]
```


# Times Series Plots
```{r}
# times series plot of the averaged S&P 500 stock returns
plot(training, type='l',
     ylab="Averaged S&P Stock Returns", main="Times Series Plot of Averaged S&P Returns by Year",
     xlab = "Year")
points(training, pch=1, cex=0.6)

log_returns <- log(training) # based on the graph, the variance was not stable
# so we will take the log transformaton 
log_returns <- ts(log_returns, frequency = 12, start = c(2015, 1)) # make sure the 
# log returns are also a times series object

plot(log_returns,type='l',
     ylab="Log of S&P 500 Stock Returns", main="Times Series Plot of Log Returns by Year",
     xlab = "Year")
points(log_returns, pch=1,cex=0.6)
# based on the graph, the variance is more stable, but the mean is not stable,
# so I will difference the date
```

# Differencing:
```{r}
# apply differencing
diff_1st_log_returns <- diff(log_returns) # first difference
ds_log_returns <- diff(diff_1st_log_returns, 12) # this is seasonal difference
diff_log_returns_seasonal <- diff(ds_log_returns) # first diff to seasonal diff

# plots
par(mfrow=c(3, 1))
plot(diff_1st_log_returns, xlab='Year', ylab='First Difference', type='l',
     main="Time Series Plot of Differenced S&P 500 Returns")
points(diff_1st_log_returns,pch=1,cex=0.7)

plot(ds_log_returns, xlab='Year', ylab='Seasonal Difference', type='l', 
     main="Times Series Plot of Seasonal Differenced S&P 500 Returns")
points(ds_log_returns, pch=1, cex=0.7)

plot(diff_log_returns_seasonal, xlab='Year', ylab='Double Differenced', type='l',
     main="Times Series Plot of Double Differenced (First and Seasonal) S&P 500 Returns")
points(diff_log_returns_seasonal, pch=1, cex=0.7)

par(mfrow=c(1, 1))
plot(diff_log_returns_seasonal, type = "l",
     ylab="Double Difference of Log S&P Returns",
     main="Double Difference of Log S&P Returns by Year",
     xlab = "Year")
points(diff_log_returns_seasonal, pch=1,cex=0.7)
# applying the log transformation, first difference, and seasonal difference makes the
# data appear stationary, we now can proceed with ARIMA model building
```

# ACF + PACF Plots for ARIMA Model Building:
```{r}
astsa::acf2(diff_log_returns_seasonal, main = "ACF Plot")

# looking at the acf, the acf is not significant at lag 1, so Q = 1 should be considered
# The ACF is significant at lag 1, but not 2 3 4, so we should consider q = 1
# the pacf is not significant at lags 12, 24, 36, etc so P = 1
# the pacf is significant at lags 1, 2, 3, and lag 11 so for now we will consider
# p = 1, 2, 3, because smaller values are better
# no exponential decay in acf plot and pacf means that there is no arma model
# d and D = 1 because we apploed differencing to both
```

# ARIMA Model Building
```{r}
diff_seasonal_log_model_1 <- sarima(log_returns, 1, 1, 0, 0, 1, 1, 12)
```
```{r}
diff_seasonal_log_model_2 <- sarima(log_returns, 2, 1, 0, 0, 1, 1, 12)
```
```{r}
diff_seasonal_log_model_3 <- sarima(log_returns, 3, 1, 0, 0, 1, 1, 12)
```
```{r}
diff_seasonal_log_model_4 <- sarima(log_returns, 0, 1, 1, 0, 1, 1, 12)
```

```{r}
par(mfrow=c(2,2))
preds1 <- astsa::sarima.for(log_returns, 14, 1, 1, 0, 0, 1, 1, 12)
preds2 <- astsa::sarima.for(log_returns, 14, 2, 1, 0, 0, 1, 1, 12)
preds3 <- astsa::sarima.for(log_returns, 14, 3, 1, 0, 0, 1, 1, 12)
preds4 <- astsa::sarima.for(log_returns, 14, 0, 1, 1, 0, 1, 1, 12)

pred_test <- log(testing)
```

# ARIMA Model Performance
```{r}
# these values came from model output
aic <- c(-3.467997, -3.472737, -3.455727, -3.473302)
aicc <- c(-3.466624, -3.469961, -3.451049, -3.471929)
bic <- c( -3.387348, -3.365205, -3.321313, -3.392653)

# calculating the MAE
mae1 <- mean(preds1$pred - pred_test)
mae2 <- mean(preds2$pred - pred_test)
mae3 <- mean(preds3$pred - pred_test)
mae4 <- mean(preds4$pred - pred_test)

mae <- c(mae1, mae2, mae3, mae4)

# calculating MSE
mse1 <- mean((preds1$pred - pred_test)^2)
mse2 <- mean((preds2$pred - pred_test)^2)
mse3 <- mean((preds3$pred - pred_test)^2)
mse4 <- mean((preds4$pred - pred_test)^2)

mse <- c(mse1, mse2, mse3, mse4)

# calculating rmse
rmse <- sqrt(mse)

data_summary <- data.frame(rbind(aic, aicc, bic, mae, mse, rmse))

# creating summary table of model performance for all of the seasonal ARIMA models

rownames(data_summary) <- c("AIC", "AICc", "BIC", "MAE", "MSE", "RMSE")

kable(data_summary, booktabs=TRUE, align = "c",
      caption = "Summary of Seasonal ARIMA (p,d,q)x(P,D,Q) Models Performance",
      col.names = c("(1, 1, 0)x(0, 1, 1)", "(2, 1, 0)x(0, 1, 1)",
                    "(3, 1, 0)x(0, 1, 1)", "(0, 1, 1)x(0, 1, 1)")) %>%
  kable_styling(latex_options=c("striped")) %>%
  kable_classic(full_width = F, html_font = "Cambria")

# based on this, the 4th model does the best at minimizing the AIC, AICc, and BIC
# and has MAE, MSE, and RMSE values that are closer to 0
```


# ACF + PACF plots + Hypothesis Testing for GARCH Model Building

```{r}
par(mfcol=c(1,2))
acf(log_returns, lag=48, main="ACF of Log Returns") 
acf(log_returns, lag=48, main="ACF of Abs Log Returns")
# these plots will tell us if the data is independent or dependent, and correlated 
# or uncorrelated

# ideally, we would want the data to be dependent and uncorrelated, but minor lower order
# serial correlations are okay

# it looks like our data is correlated (based on acf of log returns) and dependent 
# (based on acf of absolute value of log returns), but we will proceed anyways
```


```{r}
t.test(log_returns) # since the test rejects the null hypothesis
# and the confidence interval exludes 0, the test supports the
# claim that the mean of this series is not 0

res <- log_returns - mean(log_returns)

Box.test(res^2, lag = 12, type = "Ljung") # this test is used to test whether
# there are ARCH effects are present
# since we have a significant box test, there are ARCH effects present

astsa::acf2(res^2)
# using the PACF, we will we see that lag 1 and lag 6 are significant
```

# GARCH Model Building
```{r}
result1 <- fGarch::garchFit(~1+garch(1,0), data = log_returns, trace = FALSE)
summary(result1)

# model does not meet assumptions as Ljung-Box test for residuals are significant
```

```{r}
result2 <- fGarch::garchFit(~1+garch(6,0), data = log_returns, trace = FALSE)
summary(result2)
# model does not meet assumptions as Ljung-Box test for residuals are significant
```


```{r}
result3 <- fGarch::garchFit(~1+garch(1,1), data = log_returns, trace = FALSE)
summary(result3)
# model does not meet assumptions as Ljung-Box test for residuals are significant
```

# GARCH Model Performance
```{r}
preds_garch1 <- fGarch::predict(result1, n.ahead=14)[,3]
preds_garch2 <- fGarch::predict(result2, n.ahead=14)[,3]
preds_garch3 <- fGarch::predict(result3, n.ahead=14)[,3]

# these values are from the model output for GARCH models
aic_garch <- c(-0.4895506, -0.4044202, -0.4710321)
bic_garch <- c(-0.4150470, -0.2057438, -0.3716939)

# calculating MAE
mae1_garch <- mean(preds_garch1 - pred_test)
mae2_garch <- mean(preds_garch2 - pred_test)
mae3_garch <- mean(preds_garch2 - pred_test)

mae_garch <- c(mae1_garch, mae2_garch, mae3_garch)

# calculating MSE
mse1_garch <- mean((preds_garch1 - pred_test)^2)
mse2_garch <- mean((preds_garch2 - pred_test)^2)
mse3_garch <- mean((preds_garch2 - pred_test)^2)

mse_garch <- c(mse1_garch, mse2_garch, mse3_garch)

# calculating RMSE
rmse_garch <- sqrt(mse_garch)

# summarizing performance metrics for GARCH models
data_summary2 <- as.data.frame(rbind(aic_garch, bic_garch, mae_garch, mse_garch, 
                                     rmse_garch))

rownames(data_summary2) <- c("AIC", "BIC", "MAE", "MSE", "RMSE")

kable(data_summary2, booktabs=TRUE, align = "c",
      caption = "Summary of Seasonal GARCH (m,n) Models Performance",
      col.names = c("(1,0)", "(6,0)","(1,1)")) %>%
  kable_styling(latex_options=c("striped")) %>%
  kable_classic(full_width = F, html_font = "Cambria")

```

# Best Model
```{r}
# since none of the GARCH models meet the residual assumption, we cannot use
# them to predict accurately (they also have bad performance metrics)
# want the best arima model, which is a pure MA model with a pure MA
# seasonal component 

preds4 <- astsa::sarima.for(log_returns, 14, 0, 1, 1, 0, 1, 1, 12)
pred_test <- log(testing)
true_ts <- ts(pred_test, start = c(2024, 1), frequency=12)
points(true_ts, col = "blue") 
```

